---
title: "R Programming"
subtitle: "I need structure!"
author: | 
  | Brad Boehmke, Ph.D.
output: 
  ioslides_presentation:
    logo: images/uc_logo2.png
    widescreen: true
    smaller: true
    transition: 0.01
---

# Key Things to Know

## What to Remember from this Section  | Primary data structures to hold your data

- [Vectors](#7)
- [Matrices](#15)
- [Lists](#22)
- [Data frames](#32)
- <strike>Arrays</strike> (I will not discuss)

<br>

Dimensions    |  Homogenous        |  Heterogenous
------------- |  ----------------  |  ------------
1D            |  Atomic Vector     |  List
2D            |  Matrix            |  Data Frame
*n*D          |  Array             |  

## What to Remember from this Section  | After this module you should be able to:

<br>

1. Apply functions to **understand** your data structure
2. **Create** vectors, matrices, lists, and data frames
3. **Extract** elements/subset these data structures
4. Apply functions to **compare** the elements in these data structures
5. Apply functions to **summarize** the elements in these data structures


# Data Structures | Understanding key properties

## Common Functions to Learn About Your Data Structure

Operator/Function               Description             
----------------------------    ----------------       
`str()`                         provides a compact view of the data structure
`class()`                       identifies the type of data structure (i.e. matrix, data frame)
`mode()`                        identifies the nature of the elements in the data structure
`length()`                      identifies the number of elements in the data structure
`ncol()`, `nrow()`, `dim()`     identifies the number of columns and rows in the data structure
`View()`                        opens a spreadsheet-style data viewer
`head()`, `tail()`              view the first or last *n* parts of a data object
`names()`                       view the names of data structure components
----------------------------    ----------------    


# Vectors | The fundamental building block

## Vectors: creating

A linear arrangement of *common* elements

We can create vectors with 4 common operators: `:, c(), seq(),` and `rep()`:

```{r, collapse=TRUE}
# create vector with increments of 1
1:10
2.5:15

# create vector by combining non-sequential elements
c(2, 10, 4)

# sequences with specified arithmetic progression
seq(from = 0, to = 30, by = 3)
seq(from = 0, to = 30, length.out = 6)

# replicates of specified values
rep(1:4, times = 2)
rep(1:4, each = 2)
```


## Vectors: understanding their properties | Back to Our Facebook Data

Each variable in our `facebook` data is simply a vector. Let's assess the `likes` vector:

```{r, collapse=TRUE}
# read in facebook data and create a vector with the likes data
facebook <- read.delim("data/facebook.tsv")
likes <- facebook$likes

class(likes)

str(likes)

length(likes)

head(likes)

tail(likes, 20)
```


## Vectors: extracting/subsetting

**IMPORTANT:** Get to know the bracket `x[]` method

<br>

<center>
`vector[element to extract]`
</center>

<br>

```{r, collapse=TRUE}
# extract elements with bracket notation
likes[1:10]

likes[c(10, 13512, 3778, 59131)]

# extract elements based on another vector
subset_vector <- c(10, 13512, 3778, 59131)
likes[subset_vector]                    
```

<br>

We can also use comparison operators to subset...


## Vectors: comparison operators

- We can use the normal binary comparison operators: `<, >, <=, >=, ==, !=`
- Produces a logical (`TRUE`/`FALSE`) vector

```{r, collapse=TRUE}
viral <- likes >= 5000
dull <- likes == 0
normal <- likes > 0 & likes < 5000

# TRUE = 1 FALSE = 0 so we can do math with these vectors
sum(viral)
sum(dull)
sum(normal)

sum(viral)/length(likes)

# identify which observations are TRUE
which(viral)
```

## Vectors: back to extracting/subsetting

Use logical operators inside `[]` to subset data:

```{r, collapse=TRUE}
# subset data
likes[likes > 15000]

# save as new vectors with actual values rather than TRUE/FALSE response
popular <- likes[likes >= 5000]
lonely <- likes[likes == 0]

# subset based on another vector
x <- likes > 0 & likes < 5000
likes[x]
```

## Vectors: summarizing

```{r, collapse=TRUE}
summary(likes)

# include na.rm = TRUE to exclude any missing data
mean(likes, na.rm = TRUE)

median(likes)

sd(likes)

range(likes)

min(likes)

max(likes)
```


## Your Turn

Save the `friend_count` variable from the `facebook` data as a vector:

<center>
`fc <- facebook$friend_count`
</center>

<br>



<font size="6">1. What are the min, max, median, and mean number of friends?</font><br><br>
<font size="6">2. How many people have 0 friends?</font><br><br>
<font size="6">3. How many people have 1,000 or more friends?</font><br><br>
<font size="6">4. Which element contains the maximum number of friends?</font>







# Matrices | Just a 2-dimensional vector

## Matrices: creating

3 ways to create matrices; most common approach uses `matrix()` 

```{r, collapse=TRUE}
# you can specify with just nrow or ncol
# note that recycling will occur if length of matrix is not a multiple of nrows
matrix(1:12, nrow = 3, ncol = 3)
```

Two additional options to create matrices:

```{r, collapse=TRUE}
rbind(c(1:3), c(4:6))

cbind(c(1:3), c(4:6))
```


## Matrices: understanding their properties

```{r, collapse=TRUE}
# 3x4 matrix of random uniform numbers between 0-1
m <- matrix(runif(12), nrow = 3)

class(m)

str(m)

length(m)

dim(m)
```


## Matrices: extracting/subsetting

Extracting individual elements and subsetting matrices leverages brackets: 

<center>
`matrix[row, col]`
</center>

<br>

```{r, collapse=TRUE}
# extract individual elements
m[1, 3]

# subset for first two rows and all columns
m[1:2, ]

# subset for second column (simplify)
m[, 2]

# subset for second column (don't simplify)
m[, 2, drop = FALSE]
```

## Matrices: comparison operators

Can use comparison operators much like you can on vectors:

```{r, collapse=TRUE}
# comparison operators returns a matrix of logical values
m > .5

# how many elements in the matrix are greater than .5
sum(m > .5)

# which elements are greater than .5
which(m > .5)

# subset with comparison returns a vector
m[m > .5]
```


## Matrices: summarizing


```{r, collapse=TRUE}
# provides summary stats for each column of the matrix
summary(m)

# regular stat functions will be applied across all elements in the matrix
mean(m)

median(m)

# get statistic for a specific row or column
mean(m[1, ])

# get statistic for every row (MARGIN = 1) or column (MARGIN = 2)
apply(m, MARGIN = 1, FUN = mean)
```

## Your Turn

Back to our facebook friend count vector:

<center>
`fc <- facebook$friend_count`
</center>

<br>

<font size="6">1. Subset `fc` to find individuals with **more than 4,600** friends</font><br><br>
<font size="6">2. Turn this subsetted vector into a **matrix** with **10 rows**</font><br><br>
<font size="6">3. What are the **dim**ensions of this matrix</font>




# Lists | An ordered collection of objects

## Lists: creating

Lists can be confusing at first because they can hold different kinds and dimensions of data

To create a list we use `list()`

```{r, collapse=TRUE}
# list of 4 items
l1 <- list(item1 = 1:3,
           item2 = letters[1:5],
           item3 = c(T, F, T, T),
           item4 = matrix(1:9, nrow = 3))

l1
```

## Lists: understanding their properties

```{r, collapse=TRUE}
# 3x4 matrix of random uniform numbers between 0-1
class(l1)

str(l1)

length(l1)

names(l1)
```

## Lists:  extracting/subsetting

Elements of lists can be extracted using 3 approaches

<center>
preserve: `list[component]` <br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; simplify: `list[[component]]` <br>
simplify: `list$component`
</center>

<br>
```{r, collapse=TRUE}
l1["item4"]

l1[["item4"]]

l1$item4
```

## Lists:  extracting/subsetting

What's the difference?

```{r, collapse=TRUE}
class(l1["item4"])

class(l1[["item4"]])

class(l1$item4)
```

We can also extract individual elements out of list components by using...brackets!

```{r, collapse=TRUE}
# get the 2nd list component
l1[[2]]

# get the 2nd element out of the 2nd list component
l1[[2]][2]

l1$item2[2]
```

## Lists: here's what I want you to know

- Many statistical modeling results will come in the form of lists
- You need to know how to extract parts of a list to access model results

## Lists: here's what I want you to know

- Many statistical modeling results will come in the form of lists
- You need to know how to extract parts of a list to access model results

```{r, collapse=TRUE}
# here's a linear regression model
model <- lm(mpg ~ wt, data = mtcars)

summary(model)
```

## Lists: here's what I want you to know

- Luckily, `model` is simply a list of statistical results for our regression model

```{r, collapse=TRUE}
# here's a linear regression model
model <- lm(mpg ~ wt, data = mtcars)

class(model)

mode(model)

names(model)

str(model)
```

## Lists: here's what I want you to know

- Luckily, `model` is simply a list of statistical results for our regression model
- so if you want to extract the residuals or fitted values you can just use normal list subsetting procedures:

```{r, collapse=TRUE}
# extract the regression model residuals
model$residuals
```

## Your Turn

Create this linear regression model:

```{r, collapse=TRUE}
fb_model <- lm(friend_count ~ gender + age, data = facebook)
```

<br>

<font size="6">1. **Extract** the residuals from the `fb_model` list</font><br><br>
<font size="6">2. What is the **min**, **max**, **median**, and **mean** of these residuals?</font>



# Data Frames | The holy grail of data structures

## Data Frames

Data frames are R's equivalent of an Excel spreadsheet...you need to get intimate with data frames!

```{r, collapse=TRUE}
facebook
```

## Data Frames: creating

To create data frames we use `data.frame()`

```{r, collapse=TRUE}
df <- data.frame(variable.1 = 1:3,
                 variable.2 = c("a", "b", "c"),
                 variable.3 = c(TRUE, TRUE, FALSE))
```

<br>

But rarely do we create data frames from scratch; rather, we typically import them from data bases so...

<br>

...back to our `facebook` data

## Data Frames: understanding their properties


```{r, collapse=TRUE}
class(facebook)

names(facebook)

str(facebook)
```


## Data Frames: extracting/subsetting

Continue loving the brackets:

<center>
`data.frame[row, col]`
</center>

<br>
Try these: 
```{r, collapse=TRUE, eval=FALSE}
# extract the second column and all rows using column indexing or the name
facebook[, 2]
facebook[, "age"]

# extract all rows and columns 1 through 3
facebook[, 1:3]
facebook[, c("userid", "age", "dob_day")]

# subset for first row and all columns
facebook[1, ]

# use a vector to subset
columns <- c(2, 3, 5)
rows <- c(1:25)
facebook[rows, columns]
```


## Data Frames: conditional subsetting

The `subset()` function is a particularly useful function to subset data frames based on conditional statements:

<center>
`subset[data.frame, row condition, column selection]`
</center>

<br>

Try these:
```{r, collapse=TRUE, eval=FALSE}
# get all users that are 113 years old and select the first ten columns
subset(facebook, age == 113, 1:10)

# get all males older than 35 and select the likes column
subset(facebook, gender == "male" & age > 35, likes)
```

## Data Frames: summarizing

We'll cover data frame summarization in more detail in session #6

```{r, collapse=TRUE}
# summarize a single variable (column)
summary(facebook$tenure)

# summarize specific variables (columns)
summary(facebook[, c("age", "friend_count")])

# summary stats for all columns
summary(facebook)
```

## Your Turn

<font size="6">1. **Import** the `reddit` data located at<br><br> &nbsp;&nbsp;&nbsp;&nbsp;[https://bradleyboehmke.github.io/public/data/reddit.csv](https://bradleyboehmke.github.io/public/data/reddit.csv)</font><br><br><br>
<font size="6">2. What variables (column **names**) does this data frame contain?</font><br><br><br>
<font size="6">3. How many users have the **employment status** of "Student"?</font><br><br><br>
<font size="6">4. **Subset** the `reddit` data frame for only those individuals that<br>&nbsp;&nbsp;&nbsp;&nbsp;are students.</font>



# Key Things to Remember

## Remember These Functions!

Operator/Function               Description             
----------------------------    ----------------      
`str()`                         provides a compact view
`class()`                       identifies the type of data structure (i.e. matrix, data frame)
`mode()`                        identifies the nature of the elements in the data structure
`length()`                      identifies the number of elements in the data structure
`ncol()`, `nrow()`, `dim()`     identfies the number of columns and rows in the data structure
`View()`                        opens a spreadsheet-style data viewer
`head()`, `tail()`              view the first or last *n* parts of a data object
`names()`                       view the names of data structure components
`:`, `c()`, `seq()`, `rep()`    create vectors
`matrix()`                      create matrices
----------------------------    ----------------     

## Remember These Functions!

Operator/Function               Description             
----------------------------    ----------------      
`list()`                        create lists
`data.frame()`                  create data frames
`[]`, `[[]]`, `$`               extract/subset data structures
`subset()`                      conditional subsetting for data frames
`<, >, <=, >=, ==, !=`          comparison operators
`sum()`, `mean()`, `median()`   statistical summary functions
`sd()`, `range()`, `min()`      statistical summary functions
`max()`, `summary()`            statistical summary functions
----------------------------    ----------------   

## Break {.flexbox .vcenter}

<div class="centered">

<font size="10">5 minutes!</font>

</div>