---
title: "R Programming"
subtitle: "Wait, there are different types of data?"
author: | 
  | Brad Boehmke, Ph.D.
output: 
  ioslides_presentation:
    logo: images/uc_logo2.png
    widescreen: true
    smaller: true
    transition: 0.01
---


# Key Things to Remember

## What to Remember from this Section

- R has the ability to work with a variety of data *types*
- As an analyst, you should become familiar with dealing with the following types of data:

Data Type                       Description             
-----------------------         ----------------  
- [Numbers](#4)                 integer (i.e. 1,2,3...), double (i.e. 1.5, 3.66)         
- [Character Strings](#9)       "r", "I attend UC", etc.
- Regular expressions           patterns within text strings
- [Factors](#14)                nominal (male, female), ordinal (freshman, sophmore, junior), interval ($0-25, $26-50, $51-75)
- [Dates](#20)                  calendar dates (i.e. 2016-08-06, 08/06/2016), weekdays, hours, etc.
- [Logical](#28)                `TRUE`, `FALSE`, `any`, `all`
-----------------------         ---------------- 

# Numbers

## Numbers: two types of numbers

Numeric data pimarily comes in two forms: integer & double (double precision floating point)

```{r, collapse=TRUE}
# create a string of double-precision values
dbl_var <- c(1, 2.5, 4.5)  
class(dbl_var)

# placing an L after the values creates a string of integers
int_var <- c(1L, 6L, 10L)
class(int_var)
```

We can coerce integers to doubles and vice versa with `as.double()` and `as.integer()`

```{r, collapse=TRUE}
as.integer(dbl_var)

int_to_dbl <- as.double(int_var)
class(int_to_dbl)

# Combining double and integer will automatically coerce to the simplest form (double)
c(dbl_var, int_var)
```

## Numbers: comparing numbers

You've already seen logical operators using `==, !=, <, <=, >, >=`

```{r, collapse=TRUE}
x <- c(4, 4, 9, 12)
y <- c(4, 4, 9, 12.00000008)

x == y
```

Can also test for **exact equality** with `identical()` and **near equality** with `all.equal()`

```{r, collapse=TRUE}
z <- c(4, 4, 9, 12)

identical(x, y)
identical(x, z)

all.equal(x, y)
```


## Numbers: rounding

We can also round numbers multiple ways:

```{r, collapse=TRUE}
x <- c(1, 1.35, 1.7, 2.05, 2.4, 2.75)

# round to the nearest integer
round(x)

# round up
ceiling(x)

# round down
floor(x)

# round to a specified decimal
round(x, digits = 1)
```

## Your Turn

<font size="6">
Import the `numbers-your-turn.csv` file in the data folder
</font>

<br><br><br><br><br>

<font size="6">1. Are the vectors `x` & `y` equal? Exactly or approximately equal?</font><br><br>
<font size="6">2. Are the vectors `y` & `z` equal? Exactly or approximately equal?</font><br><br>
<font size="6">3. Round `x` & `y` numbers to the 4th digit</font><br><br>
<font size="6">4. Are these vectors equal now?</font>


# Characters | "Hello world!"

## Characters: creating simple character strings

Create character strings using `""`

```{r, collapse=TRUE}
a <- "learning to create"
b <- "character strings"
```

Combine character strings with `c`, `paste()` or `paste0`

```{r, collapse=TRUE}
# create a vector containing two elements - a and b
c(a, b)

# create a vector containing one element - a and b combined
paste(a, b)

# paste multiple strings
paste("I", "love", "R")

# change the separator
paste("I", "love", "R", sep = "-")

# collapse space between characters
paste0("I", "love", "R")
```


## Characters: test, conversion & coercion

Use `class()`, `mode()` and/or `is.character()` to assess the data type

```{r, collapse=TRUE}
a <- "Life of"
b <- pi

class(a)

mode(a)

is.character(pi)
```

Use `as.character()` to convert non-character to a character

```{r, collapse=TRUE}
as.character(pi)
```

Combining characters and non-characters will coerce all inputs to a character

```{r, collapse=TRUE}
c(a, b)
```

## Characters: summarizing

Use `length()` to count the number of elements (individual character strings) in a vector

```{r, collapse=TRUE}
length("How many elements are in this string?")

length(c("How", "many", "elements", "are", "in", "this", "string?"))
```

<br>

Use `nchar()` to count the number of characters in each element

```{r, collapse=TRUE}
nchar("How many characters are in this string?")

nchar(c("How", "many", "characters", "are", "in", "this", "string?"))
```

## Characters: manipulation

<center>
<blockquote>
<font size="6">
Characters are becoming a key form of data which organizations are increasingly leveraging for business analytics
</font>
</blockquote>
</center>

<br>

- Character manipulation (regex and text analysis) is a rich field; far more complex than we have time for
- More opportunities to learn
    - [Data Wrangling with R](http://uc-r.github.io/bana_datawrangling) class 
    - [Dealing with Characters](http://uc-r.github.io/characters) tutorial
    - [Dealing with Regular Expressions](http://uc-r.github.io/regex) tutorial


# Factors | aka categorical variables

## Factors: different from characters

Key Words : **finite options** and **levels**

- nominal variables
    - male, female
    - brunnette, blonde, red, black
    - Hispanic, Caucasion, Asian, African
- ordinal variables
    - slow, medium, fast
    - freshman, sophomore, junior, senior
- interval variables
    - $1-100, $101-200, $201-300
    - 0-10, 11-20, 21-30

## Factors: creating nominal factors

Create nominal factors with `factor()`

```{r, collapse=TRUE}
gender <- c("male", "female", "female")

class(gender)

gender2 <- factor(gender)

class(gender2)

gender2
```

set level preferences with `level` argument

```{r, collapse=TRUE}
factor(gender, levels = c("male", "female"))
```


## Factors: creating ordered factors

Create ordinal/interval factors with `ordered()`; set level preferences with `level` argument

```{r, collapse=TRUE}
age.range <- c("18-24", "25-34", "35-44", "45-54", "55-64", "65 or Above", "Under 18")

class(age.range)

# turn x into an ordered factor - levels default to the order of the data
age.range2 <- ordered(age.range)

class(age.range2)

age.range2
```

set level preferences with `level` argument

```{r, collapse=TRUE}
ordered(age.range, levels = c("Under 18", "18-24", "25-34", "35-44", "45-54", "55-64", "65 or Above"))
```

## Factors: summarizing

If you want to know the levels that exist in your factor variable use `levels()`

```{r, collapse=TRUE}
facebook <- read.delim("data/facebook.tsv")

levels(facebook$gender)
```

<br>

We can use the `table()` function to quickly assess the counts of each level

```{r, collapse=TRUE}
table(facebook$gender)
```

## Your Turn

<font size="6">
Import the `reddit.csv` file in the data folder
</font>

<br><br><br>

<font size="6">1. What are the levels for the `income.range` variable?</font><br><br><br>
<font size="6">2. Properly order the levels for `income.range`.</font><br><br><br>
<font size="6">3. What are the counts for each level?</font>



# Dates | The neglected variable

## Dates: creating

- The `lubridate` package makes working with dates extremely easy
- To create a date variable we simply need to know the year-month-day order

Function                        Order of elements in date-time            
-----------------------         -------------------------------  
`ymd()`                         year, month, day
`ydm()`                         year, day, month
`mdy()`                         month, day, year
`dmy()`                         day, month, year
`hm()`                          hour, minute
`hms()`                         hour, minute, second
`ymd_hms()`                     year, month, day, hour, minute, second
-----------------------         ------------------------------- 


## Dates: creating

- The `lubridate` package makes working with dates extremely easy
- To create a date variable we simply need to know the year-month-day order

```{r, collapse=TRUE, warning=FALSE, message=FALSE}
dates <- c("2015-07-01", "2015-08-01", "2015-09-01")

class(dates)
```

Convert this character string to date format with `lubridate`'s `ymd()` function

```{r, collapse=TRUE, warning=FALSE, message=FALSE}
# install.packages("lubridate") # run this line if you have not yet installed lubridate
library(lubridate)

dates2 <- ymd(dates)

class(dates2)

dates2
```


## Dates: create by merging

- Sometimes your date data are collected in separate elements 
- To convert these separate data into one date object incorporate the `ISOdate()` function:

```{r, collapse=TRUE}
yr <- c("2012", "2013", "2014", "2015") 
mo <- c("1", "5", "7", "2")
day <- c("02", "22", "15", "28")

# ISOdate converts to a POSIXct object
full_date <- ISOdate(year = yr, month = mo, day = day)
full_date
```

<br>

We can truncate the unused time data by converting with `as.Date()`
```{r, collapse=TRUE}
as.Date(full_date)
```

## Dates: extract & manipulate

We can also easily extract components of dates using `lubridate`

Function        Date-time element to extract            
------------    -------------------------------  
`year()`        Year        
`month()`       Month       
`week()`        Week        
`yday()`        Day of year       
`mday()`        Day of month   
`wday()`        Day of week    
`hour()`        Hour  
`minute()`      Minute
`second()`      Second
`tz()`          Time zone
------------    ------------------------------- 

## Dates: extract & manipulate

We can also easily extract components of dates using `lubridate`

Extract time components:

```{r, collapse=TRUE}
year(full_date)

week(full_date)

wday(full_date, label = TRUE)
```


Manipulate or change date-time components by using the function and then assignment

```{r, collapse=TRUE}
as.Date(full_date)

year(full_date) <- c(2014, 2015, 2015, 2016)

as.Date(full_date)
```


## Dates: summarizing

- We can also do regular statistical summaries of date objects
- Illustrate with the `lakers` data set that comes with the `lubridate` package

```{r, collapse=TRUE}
dates <- ymd(lakers$date)

min(dates)

max(dates)

mean(dates)

median(dates)

summary(dates)
```


## Your Turn

<font size="6">
Import the `facebook.tsv` file in the data folder
</font>

<br><br><br>

<font size="6">
1. Create a new date variable that combines the `dob_day`,<br>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`dob_month`, & `dob_year` variables.
</font><br><br><br>
<font size="6">
2. What is the `min`, `max`, `mean`, and `median` date of births in<br>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;this data frame?
</font><br><br><br><br><br>

**NOTE:** If you save the new variable as `facebook$dob <- _____________` it will add this new variable to the facebook data frame




# Logical | Boolean as in George Boole

## Logical: the basics

We already saw how we can get `TRUE`/`FALSE` responses from comparing elements
```{r, collapse=TRUE}
x <- c(4, 4, 9, 12, 2, 2, 10)
y <- c(4, 5, 9, 13, 2, 1, 10)

x == y
```

This is just vector containing logical elements
```{r, collapse=TRUE}
z <- x == y

class(z)
```

We can assess if any or all the elements are `TRUE`
```{r, collapse=TRUE}
any(z)

all(z)
```


# Key Things to Remember

## Remember These Functions!

Operator/Function                       Description             
-----------------------------           ----------------      
`as.double()`, `as.integer()`           coerce to double floating point or integer numbers
`identical()`, `all.equal()`            test for exact and near equality
`round()`, `ceiling()`, `floor()`       round numbers
`c()`, `paste()`, `paste0()`            combine character strings
`as.character()`                        coerce non-character to a character
`nchar()`                               count the number of characters in each element
`factor()`, `ordered()`                 create or coerce to factor variables
`levels()`                              assess the levels of a factor
`table()`                               get the counts of each level
-----------------------------           ----------------     

## Remember These Functions!

Operator/Function               Description             
----------------------------    ----------------      
`ymd()`, `mdy()`, `hm()`, etc   `lubridate`: create or convert to date-time variable
`Isodate()`                     create date variable by mergine separate date components
`as.Date()`                     truncate date-time variable to just date variable
`year()`, `week()`, etc         `lubridate`: extract individual date components
`any()`, `all()`                assess if any or all elements are `TRUE`
----------------------------    ----------------   

## Break {.flexbox .vcenter}

<div class="centered">

<font size="10">5 minutes!</font>

</div>