---
title: "R Programming"
subtitle: "Advanced Graphics with ggplot"
author: | 
  | Brad Boehmke, Ph.D.
output: 
  ioslides_presentation:
    logo: images/uc_logo2.png
    widescreen: true
    smaller: true
    transition: 0.01
---


# Key Things to Remember

## What to Remember from this Section

Every ggplot2 plot has three key components:

1. Data
2. A set of aesthetic mappings between variables in the data and visual
properties
3. At least one layer which describes how to render each observation. Layers
are usually created with a geom function.


# GR&A | That means grounds rules & assumptions

## R package used...

<br><br><br>

<center>
<font size="7" color="#d6d6d6">
install.packages("ggplot2")
</font>
</center>

<br>

<center>
<font size="7">
library(ggplot2)
</font>
</center>

<br><br><br><br>

<center>
<font size="6">
We'll use the built-in `mpg` data set
</font>
</center>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
```



# Fundamentals of ggplot

## First Layer is, well, boring

```{r, eval=FALSE}
# equivalent syntax
ggplot(data = mpg, aes(x = displ, y = hwy))
ggplot(mpg, aes(displ, hwy))
```


```{r, fig.align='center', echo=FALSE, fig.height=4}
ggplot(mpg, aes(displ, hwy))
```

## Let's Add Geoms

To display the data we need to tell ggplot what to draw

-----------------       ----------------------------------     
`geom_histogram()`      histogram
`geom_freqpoly()`       frequency polygon
`geom_bar()`            bar chart
`geom_point()`          scatter plot
`geom_line()`           line chart
`geom_boxplot()`        boxplot
-----------------       ----------------------------------

<br>

Check out all the available geoms at [docs.ggplot2.org/current](docs.ggplot2.org/current)

# Univariate Geoms

## Histogram & Frequency Polygons

```{r, eval=FALSE}
ggplot(data = mpg, aes(x = hwy)) +
        geom_histogram()

ggplot(data = mpg, aes(x = hwy)) +
        geom_freqpoly()
```

```{r, echo=FALSE, fig.align='center', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
p1 <- ggplot(data = mpg, aes(x = hwy)) +
        geom_histogram()

p2 <- ggplot(data = mpg, aes(x = hwy)) +
        geom_freqpoly()

library(gridExtra)
grid.arrange(p1, p2, ncol = 2)
```

## Histogram & Frequency Polygons
```{r, eval=FALSE}
ggplot(data = mpg, aes(x = hwy)) +
        geom_histogram(bins = 25, color = "grey30", fill = "white")

ggplot(data = mpg, aes(x = hwy)) +
        geom_freqpoly(bins = 25, color = "blue")
```

```{r, echo=FALSE, fig.align='center', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
p1 <- ggplot(data = mpg, aes(x = hwy)) +
        geom_histogram(bins = 25, color = "grey30", fill = "white")

p2 <- ggplot(data = mpg, aes(x = hwy)) +
        geom_freqpoly(bins = 25, color = "blue")

library(gridExtra)
grid.arrange(p1, p2, ncol = 2)
```


## Bar Chart
```{r, eval=FALSE}
ggplot(data = mpg, aes(x = class)) +
        geom_bar()

ggplot(data = mpg, aes(x = cyl)) +
        geom_bar()

ggplot(data = mpg, aes(x = factor(cyl))) +
        geom_bar()
```

```{r, echo=FALSE, fig.align='center', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
p1 <- ggplot(data = mpg, aes(x = class)) +
        geom_bar()

p2 <- ggplot(data = mpg, aes(x = cyl)) +
        geom_bar()

p3 <- ggplot(data = mpg, aes(x = factor(cyl))) +
        geom_bar()

grid.arrange(p1, p2, p3, ncol = 3)
```


## Your Turn

<br>

<font size="6">
Import the `facebook.tsv` file in the data folder
</font>

<br>

<blockquote>
<font size="6">
1. Assess the distribution of age. What does this tell you?
</font>
</blockquote>

<br>

<blockquote>
<font size="6">
2. Assess the distribution of age with a boxplot. What does this tell you?
</font>
</blockquote>

<br>

<blockquote>
<font size="6">
3. Assess the frequency of each gender. What does this tell you?
</font>
</blockquote>


## Solution

# Bivariate Geoms



# Multivariate Capabilities



