---
title: "R Programming"
subtitle: "Get me some data!"
author: | 
  | Brad Boehmke, Ph.D.
output: 
  ioslides_presentation:
    logo: images/uc_logo2.png
    widescreen: true
    smaller: true
    transition: 0.01
---


# Key Things to Know

## What to Remember from this Section
<br><br>

1. Lots of built-in data sets provided by R
2. Text files use delimiters - these determine the function to read in the data
3. There is no base R function to read in Excel data - must use a package
4. Web scraping is a dense subject but scraping online text/Excel files *can* be as easy as reading in these files from your hard drive


# Built-in Data

## Built-in Data Sets

- R has many built-in data sets
- type `data()` into your console (104 data sets should appear)

```{r}
mtcars
```

## Built in Data Sets

- R also has many convenient built-in character strings to be aware of
- Try these:

```{r, collapse=TRUE, eval=FALSE}
letters
LETTERS
month.abb
month.name
state.abb
state.division
state.name
state.region
```

<br>

### Note

- to load any of these data sets/strings into your global environment: `data("name of data set")`
- type `?name` to get more information about the built-in data


## Your Turn

<br>
<br>

<font size="10">
`1.` Load the `iris` data set

<br>

`2.` What is this data measuring?
</font>





# Importing Text Files

## Importing Text Files | .csv .txt .tsv

- Text files are a popular way to hold and exchange tabular data
- Text file formats use **delimiters** to separate the different elements
    - .csv
    - .txt
    - .tsv
- These delimiters help us know what functions to use to read in the data
- I'll illustrate with the following files in the R-Bootcamp download file
    - `mydata.csv`
    - `mydata.txt`
    - `mydata.tsv`

## Importing Text Files | Base R

Use `read.csv` for comma separated files (automatically sets the `sep = ","`)
```{r, collapse=TRUE, warning=FALSE}
read.csv("data/mydata.csv")
```
<br>

Use `read.delim` for tab delimited files (automatically sets the `sep = "\t"`)
```{r, collapse=TRUE, warning=FALSE}
read.delim("data/mydata.txt")
```

## Importing Text Files | Base R

When importing data, save to a file by using the assignment operator:

```{r, collapse=TRUE, warning=FALSE}
mydata <- read.delim("data/mydata.tsv")

mydata
```

<br>

- You now have a data object in your global environment named `mydata`
- View your data in a spreadsheet form:
    - clicking on the object in the global environment space or 
    - type `View(mydata)` in your console


## Your Turn
<br>
<br>

<font size="10">
`1.` Read in the `facebook.tsv` file
</font>

<br>

<font size="10">
`2.` Save it as an object titled `facebook`
</font>

<br>

<font size="10">
`3.` Take a peek at what this data looks like
</font>





# Importing Excel Files

## Importing Excel Files

- Excel is still the spreadsheet software of choice
- Base R does not have an Excel importing function but we can use the [`readxl`](https://cran.rstudio.com/web/packages/readxl/) package

```{r, collapse=TRUE, warning=FALSE}
# if you haven't installed the readxl package run the following line (minus hashtag)
# install.packages("readxl")

library(readxl)

read_excel("data/mydata.xlsx", sheet = "Sheet5")
```

- Benefits of `readxl`
    - recognizes date and time variables
    - keeps text variables as characters rather than factors (more to come on this)
    - does not collapse variable names
    - read more about `readxl` capabilities [here]()
- `xlsx` is an alternative package for reading in Excel files

## Your Turn
<br>

<font size="7">`1.` Read in the spreadsheet titled</font><br><br>
<font size="6">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`3. Median HH income, metro` in the 
`PEW Middle` <br>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`Class Data.xlsx` file</font><br><br><br>
<font size="7">`2.` Save it as an object titled `pew`</font><br><br><br>
<font size="7">`3.` Take a peek at what this data looks like</font>

<br>
<br>

**Hint:** You will need to skip 5 lines &#9755; check out the help file on `read_excel` to do this



# Scraping Online Files

## Scraping Online Files | Tabular files

- Dense subject with *LOTS* to learn
- We'll only focus on reading in tabular and Excel files stored online
- For more info regarding "real web scraping" see the tutorial at: [uc-r.github.io/scraping](uc-r.github.io/scraping)

<br>

Let's download some data from [https://www.data.gov/metrics](https://www.data.gov/metrics):

```{r, collapse=TRUE, eval=FALSE}
# the url for the online CSV
url <- "https://www.data.gov/media/federal-agency-participation.csv" 

# use read.csv to import
data_gov <- read.csv(url, stringsAsFactors = FALSE)

View(data_gov)
```

## Scraping Online Files | Excel files

- Scraping online Excel files follows a similar process
- `gdata` package is particular easy to use

<br>

Let's download some data from [Fair Market Rents for Section 8 Housing](http://catalog.data.gov/dataset/fair-market-rents-for-the-section-8-housing-assistance-payments-program/resource/2d33555e-5d21-4baf-84a2-e4fb3aed0807):

```{r, collapse=TRUE, eval=FALSE}
library(gdata)

url <- "http://www.huduser.org/portal/datasets/fmr/fmr2015f/FY2015F_4050_Final.xls"

# use read.xls to import
rents <- read.xls(url)

View(data_gov)
```

## Your Turn
<br>
<br>

<font size="7">
`1.` Download the file stored at:<br>
</font>

<font size="5">
`https://bradleyboehmke.github.io/public/data/reddit.csv`
</font>

<br>

<font size="7">
`2.` Save it as an object titled `reddit`
</font>

<br>

<font size="7">
`3.` Take a peek at what this data looks like
</font>



# Key Things to Remember

## Remember These Functions!

Operator/Function       Description             
-----------------       ----------------       
`data()`                access built-in data sets              
`?`                     will provide you information regarding built-in data (i.e. `?mtcars`)                
`read.csv()`            base R function for reading in .csv files (can also be used to read in a .csv file stored online)
`read.delim()`          base R function for reading in .txt and .tsv files
`read_excel()`          imports Excel data (provided by the `readxl` package)
`read.xls()`            imports Excel data stored online (provided by the `gdata` package)
`View()`                opens a spreadsheet-style data viewer
-----------------       ----------------        

## Break {.flexbox .vcenter}

<div class="centered">

<font size="10">5 minutes!</font>

</div>